---
title: "ggplot2 notes and practice"
author: "Atreya"
date: "27 February 2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Grammar of Graphics

Just like sentences, plots also have a grammar. Consider the sentence below for instance

*The quick brown fox jumps over the lazy dog*

Each word is an important element and has a role to play in this sentence

* *The*- Article
* *quick brown*- Adjective
* *fox*- Noun
* *jumps*- Verb
* *over*- Preposition
* *the* - Article
* *lazy* - Adjective
* *dog* - Noun

In a similar fashion, even a plotting framework consists of certain essential elements that make a plot meaningful as a whole

**Element**  |**Descrption**
-------------|-------------------------------------------------
*Data*       |The dataset being plotted
*Aesthetics* |The scales onto which we *map* our data
*Geometries* |The visual elements used for data
*Facets*     |Plotting small multiples
*Statistics* |Representations of our data to aid unverstanding
*Coordinates*|The space on which the data will be plotted
*Theme*      | All non-data ink

**Element**  |**Example**
-------------|------------------------------------------------------------------------------
*Data*       |Variables of interest
*Aesthetics* |x-axis, y-axis, color, fill, size, labels, alpha, shape, line width, line type
*Geometries* |point, line, histogram, bar, boxplot
*Facets*     |columns, rows
*Statistics* |binning, smoothing, descriptive, inferential
*Coordinates*|cartesian, fixed, polar, limits
*Theme*      |non-data ink

Load the standard `mtcars` dataset and check its structure
```{r warning=FALSE,message=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
data("mtcars")
str(mtcars)
head(mtcars)
```

Explanation of some variables names :

`drat` - Rear axle ratio, `disp`-Engine Displacement(in), `qsec`- 1/4 mile time, `vs`- V/S engine, `am`- Transmission (0=automatic,1=manual), `gear`- number of forward gears, `carb`- number of carburetors

Some variables need to be converted to factors before we go ahead with plotting
```{r}
mtcars %>% select(cyl) %>% n_distinct()
mtcars %>% select(vs) %>% n_distinct()
mtcars %>% select(am) %>% n_distinct()
mtcars %>% select(gear) %>% n_distinct()
mtcars %>% select(carb) %>% n_distinct()
fact_cols <- c("cyl","am","gear","carb","vs")
mtcars[fact_cols]<-lapply(mtcars[fact_cols], factor) 
```

### Horse power vs MPG :Layer by layer build of the plot
#### Step 1: Data and aesthetics elements
```{r}
p <- ggplot(mtcars,aes(x=mpg,y=hp,col=disp))
p+geom_point(alpha=0.5)
```

#### Step2: Facets
```{r}
p <- ggplot(mtcars,aes(x=mpg,y=hp,col=disp))
p+geom_point(alpha=0.5)+
  facet_grid(.~cyl)
```

####Step3: Statistics layer
```{r}
p <- ggplot(mtcars,aes(x=mpg,y=hp,col=disp))
p+geom_point(alpha=0.5)+
  facet_grid(.~cyl)+
  stat_smooth(method="lm",se=FALSE,col="red")
```

#### Step 4: Coordinates layer
```{r}
p <- ggplot(mtcars,aes(x=mpg,y=hp,col=disp))
p+geom_point(alpha=0.5)+
  facet_grid(.~cyl)+
  stat_smooth(method="lm",se=FALSE,col="red")+
  scale_x_continuous("Miles per gallon")+
  scale_y_continuous("Horse power")
```

#### Step 5: Theme layer
```{r}
p <- ggplot(mtcars,aes(x=mpg,y=hp,col=disp))
p+geom_point(alpha=0.5)+
  facet_grid(.~cyl)+
  stat_smooth(method="lm",se=FALSE,col="red")+
  scale_x_continuous("Miles per gallon",breaks = c(10,20,30))+
  labs(y="Horse power",col="Engine displacement")+
  theme(panel.background = element_blank(),
        plot.background = element_blank(),
        axis.line = element_line(colour = "black"),
        panel.grid = element_blank())
```

#### Bar plots
```{r}
bar_1 <-ggplot(mtcars,aes(x=gear,fill=am))+
        geom_bar(position = "dodge",alpha=0.8)+
        labs(x="Number of cylinders",y="Number")+
        scale_fill_manual("Transmission",labels=c("Automatic","Manual"),values=c("red","blue"))+
        theme(panel.background = element_rect(fill = "white"),panel.grid = element_blank())

bar_2 <-ggplot(mtcars,aes(x=cyl,fill=am))+
        geom_bar(position = "dodge",alpha=0.8)+
        labs(x="Number of cylinders",y="Number")+
        scale_fill_manual("Transmission",labels=c("Automatic","Manual"),values=c("red","blue"))+
        theme(panel.background = element_rect(fill = "white"),panel.grid = element_blank())
grid.arrange(bar_1,bar_2,ncol=2)
```

#### Histogram
```{r message=FALSE}
ggplot(mtcars,aes(x=mpg,fill=cyl))+
geom_histogram(binwidth = 1)+
labs(x="Miles per gallon",y="Number",fill="Number of cylinders")+
theme(panel.background=element_rect(fill="white"),panel.grid=element_blank())

ggplot(mtcars,aes(x=mpg,col=cyl))+
geom_freqpoly(lwd=1)+
labs(x="Miles per gallon",y="Number",col="Number of cylinders")+
theme(panel.background=element_rect(fill="white"),panel.grid=element_blank())
```